From DSL to HPC Component-Based Runtime: A Multi-Stencil DSL Case Study

### 2016-01-20 ARTICLES/OTHER/WG_Avalon/slides.tex
clean Helene 20/01/2016

@@

% INTRODUCTION / MOTIVATION
%-------------------------------------------------------------
\begin{frame}
\frametitle{Motivation} % HPC DSL
\begin{block}{+ Domain Specific Languages}
\begin{itemize}
\item Separation of concerns (domain/implementation)
\item Easy language for the user
\item Implicit optimizations
\item Implicit parallelization
\end{itemize}
\end{block}
\begin{alertblock}{- Domain Specific Languages}
\begin{itemize}
\item Difficulties deported to the DSL designer
\begin{itemize}
\item Low level high performance programming
\item Maintainability and portability
\end{itemize}
\item As many DSLs as domains
\begin{itemize}
\item DSL composition ?
\end{itemize}
\end{itemize}
\end{alertblock}
\end{frame}
%-------------------------------------------------------------
\begin{frame}
\frametitle{Motivation} % components
\begin{block}{Component models}
\begin{itemize}
\item Divide an application into several independent black boxes
\item Each component defines its interactions with outer world
\item Application = Assembly of components
\end{itemize}
\end{block}
\begin{block}{+ Component models}
\begin{itemize}
\item Maintainability through separation of concerns
\item Code-reuse and productivity
\item Dynamic assembly of components
\end{itemize}
\end{block}
\end{frame}
%-------------------------------------------------------------
\begin{frame}
\frametitle{Motivation} % components
What if a DSL produces a component-based runtime ?

\begin{itemize}
\item Is it feasible ?
\item Is it efficient ?
\item \textcolor{gray}{Does it improve issues of DSLs ?}
\begin{itemize}
\item \textcolor{gray}{maintainability}
\item \textcolor{gray}{portability}
\item \textcolor{gray}{productivity}
\end{itemize}
\end{itemize}

Let's take a useful example: \textit{the Multi-Stencil Language }!
\end{frame}
%-------------------------------------------------------------

@@

%-------------------------------------------------------------
\section{Multi-Stencil Language}
% related work, multi-stencil problem, language
%-------------------------------------------------------------
%----------------
\subsection{Multi-Stencil Language}
%----------------
\begin{frame}
\frametitle{Numerical simulation = Multi-Stencil application}
\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}
\def \n {5}
\def \radius {3cm}
\def \margin {8} % margin in angles, depends on the radius

@@

%----------------
\begin{frame}
\frametitle{Multi-Stencil program}
\begin{center}
$\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta,\Gamma)$
\end{center}
\begin{itemize}
\item $T$ the set of time iterations to tun the simulation
\item $\mathcal{M}$ the mesh of the simulation
\item $\mathcal{E}$ the set of mesh entities
\item $\mathcal{D}$ the set of computation domains
\item $\Delta$ the set of data
\item $\Gamma$ the set of computations
\end{itemize}
\center \textit{= the six sections of a \emph{Multi-Stencil Language} program !}
\end{frame}
%----------------

@@

\begin{lstlisting}[style=base]
@mesh:@ cart
@mesh entities:@ cell@,@edgex
@computation domains:@
  allcell @in@ cell
  alledgex @in@ edgex
@data:@
  A@,@cell
  C@,@edgex
@time:@500
@computations:@
  A@[@allcell@]@=&comp&@(@C@[@n1@]@@)@
\end{lstlisting}
\end{column}
\end{columns}
\end{frame}

@@

%----------------
\begin{frame}
\frametitle{Multi-Stencil Language}
\begin{alertblock}{MSL is not}
\begin{itemize}
\item a new stencil optimizer/compiler
\item a new distributed data structure
\end{itemize}
\end{alertblock}
\begin{block}{MSL is}
\begin{itemize}
\item a high-level language for multi-stencil simulations
\item agnostic from the type of mesh used (data structure)
\item based on identifiers only
\end{itemize}
\end{block}
\begin{center}
\textit{MSL produces a "ready-to-fill"\\component-based parallel scheduling of the simulation}
\end{center}
\end{frame}
%----------------

%----------------
\begin{frame}
\frametitle{Related Work}
\begin{block}{Complementary work}
\begin{itemize}
\item Distributed data structures: SkelGIS, Global Arrays
\item Stencil DSLs (on grids): Pochoir, PATUS
\item Stencil DSLs (on unstructured meshes): OP2, Liszt
\end{itemize}
\end{block}
\begin{alertblock}{Similar work}
\begin{itemize}
\item The SIPSim model (DDS,Data,applicators and iterators)
\begin{itemize}
\item Abstraction of a distributed data structure
\end{itemize}
\item Pipeline of stencil computations for image processing: Halide
\begin{itemize}
\item On grids (image), different abstraction level
\end{itemize}
\item DSL to component-based runtime: ?
\end{itemize}
\end{alertblock}
\end{frame}
%----------------
%-------------------------------------------------------------

%-------------------------------------------------------------
\section{Overview}
% take the same notations inside components + user
%-------------------------------------------------------------
\subsection{Overview}

%----------------
\begin{frame}
\frametitle{MSL to Component-based runtime}
\begin{block}{Ready-to-fill parallel scheduling: mid-grain parallelism}
\begin{itemize}
\item Data parallelism
\begin{itemize}
\item External distributed data structure
\item Automatic detection of synchronizations
\end{itemize}
\item Task parallelism
\begin{itemize}
\item Compile a static scheduling of computation kernels
\end{itemize}
\end{itemize}
\end{block}
The fine grain parallelism is left to other languages: 
\begin{itemize}
\item OpenMP in the kernels
\item Kernels generated by stencil compilers for CPU or GPU (Pochoir, Liszt etc.)
\end{itemize}
\end{frame}
%----------------

@@

@@

%----------------
\begin{frame}
\frametitle{Data parallelism}
\begin{enumerate}
\item Assembly of components duplicated on each resource
\item External Distributed Data Structure to split data among resources
\item Detect when synchronizations are needed
\end{enumerate}

\medskip
\begin{block}{Synchronization}
When a computation read a data, usign a stencil shape, that has been written by a previous computation.
\end{block}

@@

   \path[-]
    (C) edge[dashed] node {} (seq0);
\end{tikzpicture}
}
\end{center}
% \end{column}
% \end{columns}
\end{frame}
%----------------

@@

%----------------
\begin{frame}
\frametitle{Resume}
The MSL compiler can produce:
\begin{itemize}
\item A data parallel pattern of the multi-stencil application
\item \textcolor{gray}{An hybrid (data + task) pattern of the multi-stencil application}
\end{itemize}
\end{frame}
%----------------

@@

\medskip
\textbf{Evaluation of the data parallelism}
\begin{itemize}
\item Full SkelGIS implementation (DDS + specific interfaces to hide communications)
\item MSL implementation which uses the SkelGIS DDS
%\item Cluster Edel of Grid'5000: 2 8-cores Intel Xeon E5520, 24GB RAM / nodes, Infiniband
\item Thin Nodes TGCC Curie: two 8-cores Intel Sandy Bridge 2.7GHz, 64GB RAM, Infiniband
\end{itemize}
\end{frame}

@@

%-------------------------------------------------------------
\begin{frame}
\begin{block}{Perspectives}
\frametitle{Perspectives}
\begin{itemize}
\item Improvment of the language (convergence criteria, reduction etc.)
\item Scalability up to $32k$ cores on TGCC Curie (CEA)
\begin{itemize}
\item Compared with SkelGIS and MPI only
\end{itemize}
\item Evaluations on Data+Task parallelism
\begin{itemize}
\item OpenMP 3 inside kernels
\end{itemize}
\item Dynamic scheduling
\begin{itemize}
\item OpenMP 4 with a scheduling component (libgomp)
\item Kstar for StarPU and XKaapi runtimes
\end{itemize}
\item CPU+GPGPUs using stencil compilers (Pochoir, PATUS etc.)
\end{itemize}
$\hookrightarrow$ Show portability, maintainability introduced by components
\end{block}
\end{frame}
%-------------------------------------------------------------

>>>>

\usepackage[absolute,overlay]{textpos}
\usepackage{tcolorbox}

@@

% INTRODUCTION / MOTIVATION
%-------------------------------------------------------------
\begin{frame}
\frametitle{Motivation} % HPC DSL
\begin{block}{+ Domain Specific Languages}
\begin{itemize}
\item Separation of concerns (domain/implementation)
\item Easy language for the user
\item Implicit optimizations
\item Implicit parallelization
\end{itemize}
\end{block}
\begin{alertblock}{- Domain Specific Languages}
\begin{itemize}
\item Difficulties deported to the DSL designer
\begin{itemize}
\item Low level high performance programming
\item Maintainability and portability
\end{itemize}
\item As many DSLs as domains
\begin{itemize}
\item DSL composition ?
\end{itemize}
\end{itemize}
\end{alertblock}
\end{frame}
%-------------------------------------------------------------
\begin{frame}
\frametitle{Motivation} % components
\begin{block}{Component models}
\begin{itemize}
\item Divide an application into several independent black boxes
\item Each component defines its interactions with outer world
\item Application = Assembly of components
\end{itemize}
\end{block}
\begin{block}{+ Component models}
\begin{itemize}
\item Maintainability through separation of concerns
\item Code-reuse and productivity
\item Dynamic assembly of components
\end{itemize}
\end{block}
\end{frame}
%-------------------------------------------------------------
\begin{frame}
\frametitle{Motivation} % components
\begin{tcolorbox}[width=\textwidth,title={Motivation}]
What if a DSL produces a component-based runtime ?
\begin{itemize}
\item Is it feasible ?
\item Is it efficient ?
\item Does it improve issues of DSLs ? (maintainability etc.)
\end{itemize}
\end{tcolorbox}

\begin{block}{Contributions}
\begin{itemize}
\item The Multi-Stencil Language (MSL)
\item A first component-based back end
\item No overheads introduced
\end{itemize}
\end{block}

\end{frame}
%-------------------------------------------------------------

@@

%-------------------------------------------------------------
\section{Multi-Stencil Applications}
% related work, multi-stencil problem, language
%-------------------------------------------------------------
%----------------
\subsection{Multi-Stencil Applications}
%----------------
\begin{frame}
\frametitle{Numerical simulation = Multi-Stencil application}
\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}
\def \n {5}
\def \radius {3cm}
\def \margin {8} % margin in angles, depends on the radius

@@

%----------------
\begin{frame}
\frametitle{Multi-Stencil application}
\begin{center}
$\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta,\Gamma)$
\end{center}
\begin{itemize}
\item $T$ the set of time iterations to tun the simulation
\item $\mathcal{M}$ the mesh of the simulation
\item $\mathcal{E}$ the set of mesh entities
\item $\mathcal{D}$ the set of computation domains
\item $\Delta$ the set of data
\item $\Gamma$ the set of computations
\end{itemize}
\center \textit{= the six sections of a \emph{Multi-Stencil Language} program !}
\end{frame}
%----------------

%-------------------------------------------------------------
\section{MSL Overview}
% take the same notations inside components + user
%-------------------------------------------------------------
\subsection{MSL Overview}

@@

\medskip
\textit{\textcolor{blue}{blue} = MSL and black = user identifiers}

\medskip
\begin{lstlisting}[style=base]
@mesh:@ cart
@mesh entities:@ cell@,@edgex
@computation domains:@
  allcell @in@ cell
  alledgex @in@ edgex
@data:@
  A@,@cell
  C@,@edgex
@time:@500
@computations:@
  A@[@allcell@]@=&comp&@(@C@[@n1@]@@)@
\end{lstlisting}
\end{column}
\end{columns}
\end{frame}

%----------------
\begin{frame}[fragile]
\frametitle{Where is the code ?}
Keywords + identifiers ? That's it ? Where is the code ???
\end{frame}

%----------------
\begin{frame}[fragile]
\frametitle{Where is the code ?}
\begin{block}{Data parallelism}
Whatever the mesh is and the exact computations are:
\begin{itemize}
\item mesh and data are splitted among resources
\item synchronizations detected from \textbf{computation dependencies}
\end{itemize}
\end{block}

\begin{block}{Mid-grain task parallelism}
Task = complete computation\\
Whatever the mesh is and the exact computations are:
\begin{itemize}
\item task dependencies detected from \textbf{computation dependencies}
\end{itemize}
\end{block}

\begin{alertblock}{Fine grain optimizations and parallelizations}
Stencil compilers already exist !
\end{alertblock}

\end{frame}

%----------------
\begin{frame}[fragile]
\frametitle{Where is the code ?}
\begin{block}{Data parallelism}
Whatever the mesh is and the exact computations are:
\begin{itemize}
\item mesh and data are splitted among resources
\item synchronizations detected from computation dependencies
\end{itemize}
\end{block}

\begin{block}{Mid-grain task parallelism}
Task = complete computation\\
Whatever the mesh is and the exact computations are:
\begin{itemize}
\item task dependencies detected from computation dependencies
\end{itemize}
\end{block}

\begin{alertblock}{Fine grain optimizations and parallelizations}
Stencil compilers already exist !
\end{alertblock}

\begin{textblock*}{0.8\textwidth}(20mm,32mm)
\begin{tcolorbox}[width=\textwidth,title={MSL = computation dependencies}]
  \begin{itemize}  
  \item produces a component based parallel scheduling of the simulation
  \item it is agnostic from the mesh
  \item it is agnostic from the actual numerical code
  \end{itemize}
\end{tcolorbox} 
\end{textblock*}
% based parallel scheduling of the simulation\\ It is agnostic from the mesh\\it is agnostic from the actual numerical code

\end{frame}

@@

% %----------------
% \begin{frame}
% \frametitle{Multi-Stencil Language}
% \begin{alertblock}{MSL is not}
% \begin{itemize}
% \item a new stencil optimizer/compiler
% \item a new distributed data structure
% \end{itemize}
% \end{alertblock}
% \begin{block}{MSL is}
% \begin{itemize}
% \item a high-level language for multi-stencil simulations
% \item agnostic from the type of mesh used (data structure)
% \end{itemize}
% \end{block}
% \begin{center}
% \textit{MSL produces an empty\\component-based parallel scheduling of the simulation}
% \end{center}
% \end{frame}
% %----------------

%----------------
\begin{frame}
\frametitle{Related Work}
\begin{block}{Complementary work}
\begin{itemize}
\item Distributed data structures: SkelGIS, Global Arrays
\item Stencil DSLs (on grids): Pochoir, PATUS
\item Stencil DSLs (on unstructured meshes): OP2, Liszt
\end{itemize}
\end{block}
\begin{alertblock}{Similar work}
\begin{itemize}
\item Pipeline of stencil computations for image processing: Halide
\begin{itemize}
\item On grids (image), different abstraction level
\end{itemize}
\item DSL to component-based runtime ?
\end{itemize}
\end{alertblock}
\end{frame}
%----------------
%-------------------------------------------------------------

% %----------------
% \begin{frame}
% \frametitle{MSL to Component-based runtime}
% \begin{block}{Ready-to-fill parallel scheduling: mid-grain parallelism}
% \begin{itemize}
% \item Data parallelism
% \begin{itemize}
% \item External distributed data structure
% \item Automatic detection of synchronizations
% \end{itemize}
% \item Task parallelism
% \begin{itemize}
% \item Compile a static scheduling of computation kernels
% \end{itemize}
% \end{itemize}
% \end{block}
% The fine grain parallelism is left to other languages: 
% \begin{itemize}
% \item OpenMP in the kernels
% \item Kernels generated by stencil compilers for CPU or GPU (Pochoir, Liszt etc.)
% \end{itemize}
% \end{frame}
% %----------------

%----------------
% \begin{frame}
% \frametitle{MSL to Component-based runtime}
% \begin{block}{Empty component-based parallel scheduling of the simulation}
% \begin{itemize}
% \item Data parallelism
% \begin{itemize}
% \item Use of an external distributed data structure
% \item Automatic detection of synchronizations
% \end{itemize}
% \item Mid-grain task parallelism
% \begin{itemize}
% \item A task is a complete computation (space loop)
% \end{itemize}
% \end{itemize}
% \end{block}
% The fine grain optimizations/parallelizations are left to other languages: 
% \begin{itemize}
% \item OpenMP in the kernels
% \item Kernels generated by stencil compilers (Pochoir, Liszt etc.)
% \end{itemize}
% \end{frame}
%----------------

@@

\bigskip
\begin{tikzpicture}[remember picture]
  \node[rectangle,rounded corners=3pt,thick,inner sep=5pt,draw=red,dotted] (proc) {
    \begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
     \node[component,solid] (D) at (0,0) {$Driver$};
     \node[provide,scale=0.5,solid] (Dp) at (-1,0) {};
     \node[solid] (Ds) at (-1.5,0) {start};
     \node[use,scale=0.5,solid,right=1.5cm of D] (Du1) {};
     \node[use,scale=0.5,solid,below=1.75cm of D] (Du2) {};
     \node[use,scale=0.5,solid,below=0.8cm of Du1] (Du3) {$m$};

@@

%----------------
\begin{frame}
\frametitle{Data parallelism}
\begin{enumerate}
\item Assembly of components duplicated on each resource (SPMD)
\item External Distributed Data Structure to split data among resources
\item Detect when synchronizations are needed
\end{enumerate}

\medskip
\begin{block}{Synchronization}
When a computation read a data, using a stencil shape, that has been written by a previous computation.
\end{block}

@@

   \path[-]
    (C) edge[dashed] node {} (seq0);
\end{tikzpicture}
}
\end{center}
% \end{column}
% \end{columns}
\end{frame}
%----------------
\begin{frame}[fragile]
\frametitle{Separation of concerns}
\begin{itemize}
\item Non-computer scientist who uses the DSL
\item The numerician who writes numerical codes
\item The computer-scientists who writes
\begin{itemize}
\item parallel components
\item DDS + Data
\end{itemize}
\end{itemize}
\end{frame}

@@

%----------------
% \begin{frame}
% \frametitle{Resume}
% The MSL compiler can produce:
% \begin{itemize}
% \item A data parallel pattern of the multi-stencil application
% \item \textcolor{gray}{An hybrid (data + task) pattern of the multi-stencil application}
% \end{itemize}
% \end{frame}
%----------------

@@

\medskip
\textbf{Evaluation of the data parallelism}
\begin{itemize}
\item SkelGIS DDS + manual synchronizations and fusions
\item SkelGIS DDS + MSL
\item Thin Nodes TGCC Curie: two 8-cores Intel Sandy Bridge 2.7GHz, 64GB RAM, Infiniband
\end{itemize}
\end{frame}

@@

%-------------------------------------------------------------
\begin{frame}
\begin{block}{Perspectives}
\frametitle{Perspectives}
\begin{itemize}
\item Improvment of the language (convergence criteria, reduction etc.)
\item Scalability up to $32k$ cores on TGCC Curie (CEA)
\item Evaluations on Data+Task parallelism
\begin{itemize}
\item OpenMP 3 inside kernels
\end{itemize}
\item Dynamic scheduling
\begin{itemize}
\item OpenMP 4 with a scheduling component
\item Kstar for StarPU and XKaapi runtimes
\end{itemize}
\item CPU+GPGPUs using stencil compilers (Pochoir, PATUS etc.)
\end{itemize}
$\hookrightarrow$ Show portability, maintainability introduced by components
\end{block}
\end{frame}
%-------------------------------------------------------------

%-------------------------------------------------------------
\begin{frame}
\frametitle{Photos !}
\resizebox{0.9\textwidth}{!}{\includegraphics{./images/sc15.jpg}}
\end{frame}
%-------------------------------------------------------------
